\singlespacing \vspace{-1.6cm} \LARGE Covariation among reproductive traits in flowering plants determine interactions with floral visitors

### 2022-06-22 Manuscript/Lanuza_et_al_2021.Rmd
edits

\textbf{Corresponding author:} Jose B. Lanuza \| +34 616375981\| [barragansljose\@gmail.com](mailto:barragansljose@gmail.com){.email}

@@

Although there is an increasing number of macroecological studies that investigate plant reproductive traits [@baude2016; @munoz2016; @grossenbacher2017; @moeller2017], we still have poor understanding of how reproductive traits drive interactions with floral visitors at large ecological scales [@sargent2008; @rech2016; @salguero2016; @ruger2018]. In addition, the pollination system of a great number of plant species remains unexplored and is still unclear how specific key reproductive traits like mating or compatibility system influence plant-pollinator associations [@tur2013; @devaux2014]. Interestingly, the use of trait-based approaches [@fenster2004; @rosas2014] and trait-matching analyses [@stang2009; @bartomeus2016] has shown to be of great importance when exploring the drivers of plant-pollinator interactions. For example, plant traits can define species' network roles [e.g., specialists vs generalists;  @lazaro2013; @tur2013] and plant species that occupy reproductive trait space extremes are more likely to exhibit higher levels of specialisation and be more reliant on the trait-matching with pollinators [@junker2013; @coux2016]. Indeed, morphological matching between plants and floral visitors often determines plant-pollinator interactions, and can thus strongly influence interaction network structure [@stang2009; @ibanez2012]. Because the species' morphology can determine the species' functional role in the pollination network and the combination of traits has shown to increase the predictive power of the network interactions [@eklof2013], an interesting novel approach is to investigate how traits in the multidimensional trait space determine species interaction patterns [see @dehling2016]. Thus, by exploring the reproductive spectrum of trait variation is possible to delimit the different plant reproductive strategies and explore how these are associated with the different floral visitors.

With the recent availability of large trait databases [e.g., TRY @kattge2011; and COMPADRE @salguero2015], plant ecological strategies are being increasingly examined, and are facilitating the identification of global patterns and constraints in plant form and function [@salguero2016; @diaz2016; @bruelheide2018; @carmona2021]. However, most studies with a multitrait perspective have focused on trait correlations from the leaf [@wright2004], wood [@chave2009], or root [@laughlin2021] related traits with little or no attention given to reproductive traits [@evojtko2020; @roddy2021]. Despite the lack of an holistic view that depict reproductive trait covariation patterns, there are widely recognized reproductive trait associations between pair of traits such as the negative correlation between flower size and flower number [@sargent2007; @kettle2011], the positive association between flower size and outcrossing rate [@goodwillie2010] or the association between outcrossing rate and lifespan where short lived versus perennial species tend to have low versus high levels of outcrossing [@barrett2003; @moeller2017], respectively. Although these different trait correlations (and others) have recently allowed to progress towards a conceptual framework that integrates the different floral trait relationships [@roddy2021], we still lack empirical evidence that investigates jointly these different reproductive trait associations.

@@

We selected 28 studies from 18 different countries that constituted a total of 64 plant-pollinator networks (see Table S1 and Fig. S1). These studies recorded plant-pollinator interactions in natural systems and were selected so that we had broad geographical representation across different biological communities. Although these studies differ in sampling effort and methodology, all studies provided information about plant-pollinator interactions (weighted and non-weighted), which we used to build a database of plant species that are likely to be animal pollinated. Many of these networks are freely available either as published studies [e.g.,  @carvalheiro2014; @olesen2007; @fortuna2010] or available in online archives [e.g., The Web of Life, @fortuna2010; and Mangal, @poisot2016]. In total, our network dataset constituted 60 weighted (number of visits) and 4 unweighted (presence-absence of the interaction) networks, each sampled at a unique location and year, as well as eight meta-webs where interactions were pooled across several locations and multiple years.

@@

We selected a total of 19 different functional traits that comprised reproductive traits and plant form and size related traits (see Table 1). From these, 16 were reproductive traits that consisted of 13 floral traits and 3 reproductive biology traits. Floral traits included traits related to the size of floral organs (e.g., style length), floral display (e.g., number of flowers) and floral rewards (e.g., pollen quantity). Reproductive biology traits indicated the reproductive system of the plant and included breeding, mating and compatibility system. The 3 remaining traits were plant size and form related traits that are commonly used to characterize the fast-slow continuum of plant trait variation (i.e., plant height, lifespan and life form). For each plant species, we undertook an extensive literature and online search for all traits across a wide range of resources (plant databases, online floras, books, journals and images). From a total of 30,120 possible cells considering all traits and plant species (20 columns × 1,506 species), we were able to fill 24,341 cells (80.8% of the dataset, see Fig. S2 for missing values information for each trait). An extended description of each trait and how it was obtained can be found in Appendix S1.

@@

kable(d, booktabs = T, linesep = "\\addlinespace", escape = FALSE,caption = "\\textbf{Table 1 | Quantitative and categorical traits used in this study.}") %>%
  kable_styling(full_width = F,font_size = 10) %>% row_spec(0, bold=T) %>% pack_rows("Quantitative", 1, 12) %>%
pack_rows("Categorical", 13, 19)

@@

To avoid the loss of relevant ecological information, we imputed trait missing values with the help of the function *missForest* [@stekhoven2012] which allows imputation of datasets with continuous and categorical variables. We accounted for the phylogenetic distance among species on the imputation process by including the eigenvectors of a principal component analysis of the phylogenetic distance (PCoA) which has been shown to improve the performance of *missForest* [@penone2014]. To extract the eigenvectors, we used the function *PVRdecomp* from the package *PVR* [@santos2018] based on a previous conceptual framework that considers phylogenetic eigenvectors [@diniz-filho2012]. We conducted two different imputations, one for the full set of species (1,506 species, 5.79% of missing values) excluding nectar and pollen traits because of the high percentage of missing values (Fig. S2) and a second one for the subset of species with data for pollen per flower and microliters of nectar (755 species, 8.01% of missing values). To corroborate that our imputation of missing values did not affect our results, we evaluated the reproductive spectrum (see section below) with and without missing values and we found consistent similar results on both analyses (Fig S3 and Fig S4).

@@

We used Bayesian modelling (see below for details) to explore the effect of floral visitor groups and the main axes of trait variation (pPCA with imputed dataset) on both qualitative and quantitative floral interactions per plant species. For this, we divided floral visitors into six main guilds that differ in life form, behaviour and are likely to play a similar ecological role: (i) bees (Hymenoptera-Anthophila), (ii) non-bee-Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Diptera-Syrphidae), (iv) non-syrphid-Diptera (Diptera-non-Syrphidae), (v) Lepidoptera and (vi) Coleoptera. Moreover, because the guild of bees was the most represented group with 2,256 records and had the highest frequency of visits, we also explored presence-absence of the interaction and number of visits of the main bee families (Andrenidae, Apidae, Colletidae, Halictidae and Megachilidae) on the trait space. In addition, we found that *Apis mellifera* was the floral visitor with the largest proportion of records counted (7.55% of the total). This finding is consistent with previous research showing that *A. mellifera* was the most frequent floral visitor in a similar dataset of 80 plant-pollinator networks in natural ecosystems [@hung2018]. Hence, to control for the effect of *A. mellifera* on the observed visitation patterns of bees, we conducted an analogous analysis with presence-absence of the interaction and number of visits excluding *A. mellifera*. We found that *A. mellifera*, was partly driving some of the observed trends on PC1 (Fig. S5). However, we did not detect major differences on PC2 and PC3.

@@

The phylogenetically informed principal component analysis captured by the first two and three axes 51.8% and 70.97% of the reproductive trait variation, respectively (Fig. 1 and Fig. S7). The first principal component (PC1 or the flower number - flower size axis) represented 26.72% of the trait variation and indicated a negative correlation between flower number and flower size. The main contributing traits to PC1 were plant height, flower number, ovule number and flower size (loadings \> \|0.5\|; Table S3) but style length also contributed moderately to PC1 (loading = -0.33). One end of this axis comprised species with high investment in flower number and plant height but small flower size, short style length and low ovule number. For instance, on this end of the spectrum we find the species *Cornus florida* which has approximately a total of 10.000 flowers, an average height of 7.5m, flowers of 3mm wide, a style length of 3.5mm and a total of 2 ovules per flower. The other end of this spectrum comprised species that were short in height and invested in large flowers, long styles, many ovules, but few flowers. For instance, on this side of the axis we find the species *Petunia axillaris* that has approximately 10 flowers per plant, a height of 0.5m, flowers over 50mm wide, styles of 25mm and over 200 ovules per flower. The second principal component (PC2 or the pollinator dependence axis) represented 25.05% of the trait variation and indicated the variation from low to high autonomous selfing, or in other words, high to low pollinator dependence, respectively. The main driver of trait variation on PC2 was autonomous selfing (loading = 0.85) but the other traits (except ovule number) also made moderate contributions (loadings from 0.27 to 0.4; Table S3). In general terms, species with high pollinator dependence were associated with larger and a higher number of flowers, greater plant height and longer styles. On this extreme of the spectrum we find the species *Zuccagnia punctata* that is a self-incompatible shrub that depends completely on floral visitors for seed production, it has approximately 1500 flowers per plant, 3m height and a style length of 20mm. In contrast, species with low pollinator dependence tended to have fewer and smaller flowers, shorter plant height and shorter styles. For instance, on this end of the spectrum we find the species *Veronica peregrina* which is a self-compatible herb that is thought to be almost a complete selfer that relies none or little on floral visitors, it has around 20 flowers per plant, a height of 0.2m and a style length of 0.25mm. Further, the third principal component (PC3 or the style length - pollinator dependence axis) explained a considerable amount of trait variability (19.17%) and represented a negative correlation between style length (loading = -0.66) and pollinator dependence (autonomous selfing loading = |-0.51|) where species with short styles had high pollinator dependence and species with long styles low pollinator dependence. The remaining traits, apart from ovule number, were positively correlated with style length and negatively correlated with pollinator dependence (loadings from -0.23 to -0.46; Table S3). In addition, the pPCA with the subset of species that had nectar and pollen quantity data showed that nectar quantity (microlitres of nectar per flower) and pollen grains per flower were positively associated with flower size, style length and ovule number but negatively associated with flower number (Fig. S4). This pPCA explained similar variance with the first two principal components (45.52%) and similar associations of traits despite some variability in the loadings (Table S4).

```{r, echo=FALSE, message=FALSE, cache=FALSE, results='hide', warning=FALSE, out.width="95%", fig.height=10,fig.align='center',fig.width=10,fig.cap= "\\textbf{Figure 1} Phylogenetically informed principal component analysis (pPCA) of 1,236 plant species from 28 plant-pollinator network studies. The solid arrows indicate the direction of the different quantitative traits (flower number, plant height, style length, flower size, ovule number and level of autonomous selfing) across the two main axes of trait variation. The length of the arrows indicate the weight of the variables on each principal component and the dashed lines show the opposed direction of trait variation. The icons at both ends of arrows and dashed lines illustrate the extreme form of the trait continuum."}

@@

We found (with an Anova) that most categorical traits were statistically associated with the first two axes of trait variation (Fig. 2 and Table S2). Flower symmetry, which was only associated with PC2 (Sum of squares = 8.51, F-value = 14.72, *P* \< 0.01 ), and nectar provision, which was independent of PC1 and PC2 (PC1: Sum of squares = 0.37, F-value = 0.29 , *P* = 0.59; PC2: Sum of squares = 0.83, F-value = 1.43, *P* = 0.23) showed lack of statistical association. In addition, we found (with a Tukey test) statistical differences between the different levels of categorical traits in the trait space (Fig. S8). Regarding self-compatibility, we found larger differences on PC2 (i.e., species with unisexual flowers that were self-incompatible were statistically differentiated from species with partial or full self-compatibility; Fig. S8a and Fig. S8b). Life forms differed statistically across both axes of trait variation and followed a gradient of larger life forms (trees and shrubs) with higher pollinator dependence to smaller ones (herbs) with lower pollinator dependence (Fig. S8c and Fig. S8d). Consequently, lifespan also followed this gradient but perennial and short lived species only differed statistically on PC2 (Fig. S8e and Fig. S8f). Species with unisexual flowers (monoecious and dioecious) were clustered on both extremes of the first two principal components and had the highest pollinator dependence and highest number of flowers (Fig. S8g and Fig. S8h). Moreover, we found that the campanulate and capitulum flower shapes were differentiated from tube, papilionaceous, open and brush shapes in the trait space. The former morphologies had larger flowers and greater pollinator dependence, while the latter had higher flower number and lower pollinator dependence (Fig. S8i and Fig. S8j). Regarding flower symmetry, zygomorphic flowers were associated with lower levels of pollinator dependence, whereas actinomorphic flowers had higher levels of pollinator dependence (Fig. S8k and Fig. S8l).

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, results='hide', warning=FALSE, fig.height=12, fig.width=18, out.width="80%",fig.align='center',fig.cap= "\\textbf{Figure 2} Location of the different qualitative traits on the trait space. The panel is composed by the traits that showed statistical association with the first two axes of trait variation: compatibility system (a), life form (b), lifespan (c), breeding system (d), flower shape (e) and flower symmetry (f). The solid arrows indicate the direction of variation of the different quantitative traits showed in figure 1: flower number (FN), plant height (PH), style length (SL), flower size (FS), ovule number (ON) and the level of autonomous selfing (S)."}

@@

We found a strong phylogenetic signal (*P* \< 0.01) for most quantitative traits (Table S5). The traits that showed the highest phylogenetic signal were ovule number ($\lambda$ = 1), pollen grains per flower ($\lambda$ = 1) and plant height ($\lambda$ = 0.96), followed by flower length ($\lambda$ = 0.75), flower width ($\lambda$ = 0.73), number of flowers per plant ($\lambda$ = 0.69) and nectar concentration ($\lambda$ = 0.65). The traits that showed a moderate phylogenetic signal were inflorescence width ($\lambda$ = 0.57), style length ($\lambda$ = 0.49) and autonomous selfing ($\lambda$ = 0.34). Finally, microliters of nectar per flower showed the lowest phylogenetic signal of all traits ($\lambda$ = 0.14).

@@

The main axes of trait variation (PC1, PC2 and PC3) explained partly presence-absence of interaction partners (conditional $R^{2}$ = 0.26; marginal $R^{2}$ = 0.20) but little of the overall number of visits (conditional $R^{2}$ = 0.31; marginal $R^{2}$ = 0.06). However, we found relevant differences across the different floral visitor guilds on both presence-absence of interactions and number of visits (Fig. 3). We found on the flower number - flower size axis that plants with high flower number and small flowers had higher interaction partners of Coleoptera, non-bee-Hymenoptera and all Diptera guilds (Fig. 3a) but plants with low flower number but large flowers had higher interaction partners of bees and Lepidoptera guilds. Regarding visitation rates on the flower number - flower size axis, we found similar trends but bees and syrphids guilds in this case showed higher number of visits to plant species with high flower number but small (Fig. 3d). On the pollinator dependence axis, we found that all plant species with higher pollinator dependence had higher number of interacting partners and number of visits from all floral visitor guilds (Fig. 3b and Fig. 3e). Finally, on the style length - pollinator dependence axis plant species with short styles and high pollinator dependence had higher interaction partners of all guilds but bees, that interacted clearly more with plant species with long styles and low pollinator dependence (Fig. 3c). However, regarding number of visits on the style length - pollinator dependence axis, we found that plants with long styles and high selfing interacted more frequently with Lepidoptera and non-bee-Hymenoptera guilds (Fig. 3f).

The additional model for both presence-absence of interaction (marginal $R^{2}$ = 0.29; conditional $R^{2}$ = 0.19) and number of visits (marginal $R^{2}$ = 0.30; conditional $R^{2}$ = 0.03) for the most represented families of bees showed that the family Apidae was the main driver of the observed patterns. The contrasting differences between presence-absence of interaction and number of visits for bees on PC1 (Fig. 3a and Fig. 3d) were driven by the family Andrenidae that had higher number of interacting partners but lower number of visits on plant species with low number of large flowers (Fig. S9).

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, results='hide', warning=FALSE, fig.height=14, fig.width=27,fig.cap= "**Figure 3** Fitted posterior estimates of the presence-absence of interaction (a, b and c) and number of visits (d, e and f) of the different floral visitor guilds in relation to the main axes of trait variation (PC1, PC2 and PC3). PC1 represents the flower number - flower size axis, PC2 represents the pollinator dependence axis and PC3, the style length - pollinator dependence axis. For visualization purposes, due to large differences between number of visits of bees and the rest of guilds, the number of visits was log-transformed (Y-axis of lower panel)."}

@@

The variance of the different plant species-level network metrics was poorly explained by the three main axes of trait variation (Fig. S10; number of visits \~ PCs, conditional $R^{2}$ = 0.11, marginal $R^{2}$ = 0.02; normalized degree \~ PCs, conditional $R^{2}$ = 0.24, marginal $R^{2}$ = 0.02; and, specialization \~ PCs, conditional $R^{2}$ = 0.37, marginal$R^{2}$ = 0.03). Overall, the most notable trends were found on the flower number - flower size axis (PC1) and the style length - pollinator dependence axis (PC3) for number of visits and specialization. On PC1, number of visits was higher for plant species with more flowers but was lower for plant species with larger flowers (Fig. S10a). However, specialization showed the opposite trend on this axis (Fig. S10g). On PC3, the number of visits was lower for plants with shorter styles and higher pollinator dependence but higher for plant species with longer styles and lower pollinator dependence (Fig. S10c). Again, specialization showed the opposite trend to number of visits (Fig. S10i).

>>>>

\textbf{Keywords:} life-history strategies \| plant reproductive traits \| plant-pollinator interactions

\textbf{Author for correspondence:} Jose B. Lanuza \| [barragansljose\@gmail.com](mailto:barragansljose@gmail.com){.email}

@@

Although there is an increasing number of macroecological studies that investigate plant reproductive traits [@baude2016; @munoz2016; @grossenbacher2017; @moeller2017], we still have poor understanding of how reproductive traits drive interactions with floral visitors at large ecological scales [@sargent2008; @rech2016; @salguero2016; @ruger2018]. In addition, the pollination system of a great number of plant species remains unexplored and is still unclear how specific key reproductive traits like mating or compatibility system influence plant-pollinator associations [@tur2013; @devaux2014]. Interestingly, the use of trait-based approaches [@fenster2004; @rosas2014] and trait-matching analyses [@stang2009; @bartomeus2016] has shown to be of great importance when exploring the drivers of plant-pollinator interactions. For example, plant traits can define species' network roles [e.g., specialists vs generalists @tur2013] and plant species that occupy reproductive trait space extremes are more likely to exhibit higher levels of specialisation and be more reliant on the trait-matching with pollinators [@junker2013; @coux2016]. Indeed, morphological matching between plants and floral visitors often determines plant-pollinator interactions, and can thus strongly influence interaction network structure [@stang2009; @ibanez2012]. Because the species' morphology can determine the species' functional role in the pollination network and the combination of traits has shown to increase the predictive power of the network interactions [@eklof2013], an interesting novel approach is to investigate how traits in the multidimensional trait space determine species interaction patterns [see @dehling2016]. Thus, by exploring the reproductive spectrum of trait variation is possible to delimit the different plant reproductive strategies and explore how these are associated with the different floral visitors.

With the recent availability of large trait databases [e.g., TRY @kattge2011;\ COMPADRE @salguero2015], plant ecological strategies are being increasingly examined, and are facilitating the identification of global patterns and constraints in plant form and function [@salguero2016; @diaz2016; @bruelheide2018; @carmona2021]. However, most studies with a multitrait perspective have focused on trait correlations from the leaf [@wright2004], wood [@chave2009], or root [@laughlin2021] related traits with little or no attention given to reproductive traits [@evojtko2020; @roddy2021]. Despite the lack of an holistic view that depict reproductive trait covariation patterns, there are widely recognized reproductive trait associations between pair of traits such as the negative correlation between flower size and flower number [@sargent2007; @kettle2011], the positive association between flower size and outcrossing rate [@goodwillie2010] or the association between outcrossing rate and lifespan where short lived versus perennial species tend to have low versus high levels of outcrossing [@barrett2003; @moeller2017], respectively. Although these different trait correlations (and others) have recently allowed to progress towards a conceptual framework that integrates the different floral trait relationships [@roddy2021], we still lack empirical evidence that investigates jointly these different reproductive trait associations.

@@

We selected 28 studies from 18 different countries that constituted a total of 64 plant-pollinator networks (see table S1 and figure S1). These studies recorded plant-pollinator interactions in natural systems and were selected so that we had broad geographical representation across different biological communities. Although these studies differ in sampling effort and methodology, all studies provided information about plant-pollinator interactions (weighted and non-weighted), which we used to build a database of plant species that are likely to be animal pollinated. Many of these networks are freely available either as published studies [e.g., @carvalheiro2014; @olesen2007; @fortuna2010] or available in online archives [e.g., Mangal @poisot2016;\ The Web of Life @fortuna2010]. In total, our network dataset constituted 60 weighted (number of visits) and 4 unweighted (presence-absence of the interaction) networks, each sampled at a unique location and year, as well as eight meta-webs where interactions were pooled across several locations and multiple years.

@@

We selected a total of 19 different functional traits that comprised reproductive traits and plant form and size related traits (see table 1). From these, 16 were reproductive traits that consisted of 13 floral traits and 3 reproductive biology traits. Floral traits included traits related to the size of floral organs (e.g., style length), floral display (e.g., number of flowers) and floral rewards (e.g., pollen quantity). Reproductive biology traits indicated the reproductive system of the plant and included breeding, mating and compatibility system. The 3 remaining traits were plant size and form related traits that are commonly used to characterize the fast-slow continuum of plant trait variation (i.e., plant height, lifespan and life form). For each plant species, we undertook an extensive literature and online search for all traits across a wide range of resources (plant databases, online floras, books, journals and images). From a total of 30,120 possible cells considering all traits and plant species (20 columns × 1,506 species), we were able to fill 24,341 cells (80.8% of the dataset, see figure S2 for missing values information for each trait). An extended description of each trait and how it was obtained can be found in Appendix S1.

@@

kable(d, booktabs = T, linesep = "\\addlinespace", escape = FALSE,caption = "\\textbf{Table 1.} Quantitative and categorical traits used in this study.") %>%
  kable_styling(full_width = F,font_size = 12) %>% row_spec(0, bold=T) %>% pack_rows("Quantitative", 1, 12) %>%
pack_rows("Categorical", 13, 19)

@@

To avoid the loss of relevant ecological information, we imputed trait missing values with the help of the function *missForest* [@stekhoven2012] which allows imputation of datasets with continuous and categorical variables. We accounted for the phylogenetic distance among species on the imputation process by including the eigenvectors of a principal component analysis of the phylogenetic distance (PCoA) which has been shown to improve the performance of *missForest* [@penone2014]. To extract the eigenvectors, we used the function *PVRdecomp* from the package *PVR* [@santos2018] based on a previous conceptual framework that considers phylogenetic eigenvectors [@diniz-filho2012]. We conducted two different imputations, one for the full set of species (1,506 species, 5.79% of missing values) excluding nectar and pollen traits because of the high percentage of missing values (figure S2) and a second one for the subset of species with data for pollen per flower and microliters of nectar (755 species, 8.01% of missing values). To corroborate that our imputation of missing values did not affect our results, we evaluated the reproductive spectrum (see section below) with and without missing values and we found consistent similar results on both analyses (Fig S3 and Fig S4).

@@

We used Bayesian modelling (see below for details) to explore the effect of floral visitor groups and the main axes of trait variation (pPCA with imputed dataset) on both qualitative and quantitative floral interactions per plant species. For this, we divided floral visitors into six main guilds that differ in life form, behaviour and are likely to play a similar ecological role: (i) bees (Hymenoptera-Anthophila), (ii) non-bee-Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Diptera-Syrphidae), (iv) non-syrphid-Diptera (Diptera-non-Syrphidae), (v) Lepidoptera and (vi) Coleoptera. Moreover, because the guild of bees was the most represented group with 2,256 records and had the highest frequency of visits, we also explored presence-absence of the interaction and number of visits of the main bee families (Andrenidae, Apidae, Colletidae, Halictidae and Megachilidae) on the trait space. In addition, we found that *Apis mellifera* was the floral visitor with the largest proportion of records counted (7.55% of the total). This finding is consistent with previous research showing that *A. mellifera* was the most frequent floral visitor in a similar dataset of 80 plant-pollinator networks in natural ecosystems [@hung2018]. Hence, to control for the effect of *A. mellifera* on the observed visitation patterns of bees, we conducted an analogous analysis with presence-absence of the interaction and number of visits excluding *A. mellifera*. We found that *A. mellifera*, was partly driving some of the observed trends on PC1 (figure S5). However, we did not detect major differences on PC2 and PC3.

@@

The phylogenetically informed principal component analysis captured by the first two and three axes 51.8% and 70.97% of the reproductive trait variation, respectively (figure 1 and figure S7). The first principal component (PC1 or the flower number - flower size axis) represented 26.72% of the trait variation and indicated a negative correlation between flower number and flower size. The main contributing traits to PC1 were plant height, flower number, ovule number and flower size (loadings \> \|0.5\|; table S3) but style length also contributed moderately to PC1 (loading = -0.33). One end of this axis comprised species with high investment in flower number and plant height but small flower size, short style length and low ovule number. For instance, on this end of the spectrum we find the species *Cornus florida* which has approximately a total of 10.000 flowers, an average height of 7.5m, flowers of 3mm wide, a style length of 3.5mm and a total of 2 ovules per flower. The other end of this spectrum comprised species that were short in height and invested in large flowers, long styles, many ovules, but few flowers. For instance, on this side of the axis we find the species *Petunia axillaris* that has approximately 10 flowers per plant, a height of 0.5m, flowers over 50mm wide, styles of 25mm and over 200 ovules per flower. The second principal component (PC2 or the pollinator dependence axis) represented 25.05% of the trait variation and indicated the variation from low to high autonomous selfing, or in other words, high to low pollinator dependence, respectively. The main driver of trait variation on PC2 was autonomous selfing (loading = 0.85) but the other traits (except ovule number) also made moderate contributions (loadings from 0.27 to 0.4; table S3). In general terms, species with high pollinator dependence were associated with larger and a higher number of flowers, greater plant height and longer styles. On this extreme of the spectrum we find the species *Zuccagnia punctata* that is a self-incompatible shrub that depends completely on floral visitors for seed production, it has approximately 1500 flowers per plant, 3m height and a style length of 20mm. In contrast, species with low pollinator dependence tended to have fewer and smaller flowers, shorter plant height and shorter styles. For instance, on this end of the spectrum we find the species *Veronica peregrina* which is a self-compatible herb that is thought to be almost a complete selfer that relies none or little on floral visitors, it has around 20 flowers per plant, a height of 0.2m and a style length of 0.25mm. Further, the third principal component (PC3 or the style length - pollinator dependence axis) explained a considerable amount of trait variability (19.17%) and represented a negative correlation between style length (loading = -0.66) and pollinator dependence (autonomous selfing loading = |-0.51|) where species with short styles had high pollinator dependence and species with long styles low pollinator dependence. The remaining traits, apart from ovule number, were positively correlated with style length and negatively correlated with pollinator dependence (loadings from -0.23 to -0.46; table S3). In addition, the pPCA with the subset of species that had nectar and pollen quantity data showed that nectar quantity (microlitres of nectar per flower) and pollen grains per flower were positively associated with flower size, style length and ovule number but negatively associated with flower number (figure S4). This pPCA explained similar variance with the first two principal components (45.52%) and similar associations of traits despite some variability in the loadings (table S4).

```{r, echo=FALSE, message=FALSE, cache=FALSE, results='hide', warning=FALSE, out.width="95%", fig.height=10,fig.align='center',fig.width=10,fig.cap= "\\textbf{Figure 1.} Phylogenetically informed principal component analysis (pPCA) of 1,236 plant species from 28 plant-pollinator network studies. The solid arrows indicate the direction of the different quantitative traits (flower number, plant height, style length, flower size, ovule number and level of autonomous selfing) across the two main axes of trait variation. The length of the arrows indicate the weight of the variables on each principal component and the dashed lines show the opposed direction of trait variation. The icons at both ends of arrows and dashed lines illustrate the extreme form of the trait continuum."}

@@

We found (with an Anova) that most categorical traits were statistically associated with the first two axes of trait variation (figure 2 and table S2). Flower symmetry, which was only associated with PC2 (Sum of squares = 8.51, F-value = 14.72, *P* \< 0.01 ), and nectar provision, which was independent of PC1 and PC2 (PC1: Sum of squares = 0.37, F-value = 0.29 , *P* = 0.59; PC2: Sum of squares = 0.83, F-value = 1.43, *P* = 0.23) showed lack of statistical association. In addition, we found (with a Tukey test) statistical differences between the different levels of categorical traits in the trait space (figure S8). Regarding self-compatibility, we found larger differences on PC2 (i.e., species with unisexual flowers that were self-incompatible were statistically differentiated from species with partial or full self-compatibility; figure S8a and figure S8b). Life forms differed statistically across both axes of trait variation and followed a gradient of larger life forms (trees and shrubs) with higher pollinator dependence to smaller ones (herbs) with lower pollinator dependence (figure S8c and figure S8d). Consequently, lifespan also followed this gradient but perennial and short lived species only differed statistically on PC2 (figure S8e and figure S8f). Species with unisexual flowers (monoecious and dioecious) were clustered on both extremes of the first two principal components and had the highest pollinator dependence and highest number of flowers (figure S8g and figure S8h). Moreover, we found that the campanulate and capitulum flower shapes were differentiated from tube, papilionaceous, open and brush shapes in the trait space. The former morphologies had larger flowers and greater pollinator dependence, while the latter had higher flower number and lower pollinator dependence (figure S8i and figure S8j). Regarding flower symmetry, zygomorphic flowers were associated with lower levels of pollinator dependence, whereas actinomorphic flowers had higher levels of pollinator dependence (figure S8k and figure S8l).

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, results='hide', warning=FALSE, fig.height=12, fig.width=18, out.width="80%",fig.align='center',fig.cap= "\\textbf{Figure 2.} Location of the different qualitative traits on the trait space. The panel is composed by the traits that showed statistical association with the first two axes of trait variation: compatibility system (a), life form (b), lifespan (c), breeding system (d), flower shape (e) and flower symmetry (f). The solid arrows indicate the direction of variation of the different quantitative traits showed in figure 1: flower number (FN), plant height (PH), style length (SL), flower size (FS), ovule number (ON) and the level of autonomous selfing (S)."}

@@

We found a strong phylogenetic signal (*P* \< 0.01) for most quantitative traits (table S5). The traits that showed the highest phylogenetic signal were ovule number ($\lambda$ = 1), pollen grains per flower ($\lambda$ = 1) and plant height ($\lambda$ = 0.96), followed by flower length ($\lambda$ = 0.75), flower width ($\lambda$ = 0.73), number of flowers per plant ($\lambda$ = 0.69) and nectar concentration ($\lambda$ = 0.65). The traits that showed a moderate phylogenetic signal were inflorescence width ($\lambda$ = 0.57), style length ($\lambda$ = 0.49) and autonomous selfing ($\lambda$ = 0.34). Finally, microliters of nectar per flower showed the lowest phylogenetic signal of all traits ($\lambda$ = 0.14).

@@

The main axes of trait variation (PC1, PC2 and PC3) explained partly presence-absence of interaction partners (conditional $R^{2}$ = 0.26; marginal $R^{2}$ = 0.20) but little of the overall number of visits (conditional $R^{2}$ = 0.31; marginal $R^{2}$ = 0.06). However, we found relevant differences across the different floral visitor guilds on both presence-absence of interactions and number of visits (figure 3). We found on the flower number - flower size axis that plants with high flower number and small flowers had higher interaction partners of Coleoptera, non-bee-Hymenoptera and all Diptera guilds (figure 3a) but plants with low flower number but large flowers had higher interaction partners of bees and Lepidoptera guilds. Regarding visitation rates on the flower number - flower size axis, we found similar trends but bees and syrphids guilds in this case showed higher number of visits to plant species with high flower number but small (figure 3d). On the pollinator dependence axis, we found that all plant species with higher pollinator dependence had higher number of interacting partners and number of visits from all floral visitor guilds (figure 3b and figure 3e). Finally, on the style length - pollinator dependence axis plant species with short styles and high pollinator dependence had higher interaction partners of all guilds but bees, that interacted clearly more with plant species with long styles and low pollinator dependence (figure 3c). However, regarding number of visits on the style length - pollinator dependence axis, we found that plants with long styles and high selfing interacted more frequently with Lepidoptera and non-bee-Hymenoptera guilds (figure 3f).

The additional model for both presence-absence of interaction (marginal $R^{2}$ = 0.29; conditional $R^{2}$ = 0.19) and number of visits (marginal $R^{2}$ = 0.30; conditional $R^{2}$ = 0.03) for the most represented families of bees showed that the family Apidae was the main driver of the observed patterns. The contrasting differences between presence-absence of interaction and number of visits for bees on PC1 (figure 3a and figure 3d) were driven by the family Andrenidae that had higher number of interacting partners but lower number of visits on plant species with low number of large flowers (figure S9).

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, results='hide', warning=FALSE, fig.height=14, fig.width=27,fig.cap= "**Figure 3.** Fitted posterior estimates of the presence-absence of interaction (a, b and c) and number of visits (d, e and f) of the different floral visitor guilds in relation to the main axes of trait variation (PC1, PC2 and PC3). PC1 represents the flower number - flower size axis, PC2 represents the pollinator dependence axis and PC3, the style length - pollinator dependence axis. For visualization purposes, due to large differences between number of visits of bees and the rest of guilds, the number of visits was log-transformed (Y-axis of lower panel)."}

@@

The variance of the different plant species-level network metrics was poorly explained by the three main axes of trait variation (figure S10; number of visits \~ PCs, conditional $R^{2}$ = 0.11, marginal $R^{2}$ = 0.02; normalized degree \~ PCs, conditional $R^{2}$ = 0.24, marginal $R^{2}$ = 0.02; and, specialization \~ PCs, conditional $R^{2}$ = 0.37, marginal$R^{2}$ = 0.03). Overall, the most notable trends were found on the flower number - flower size axis (PC1) and the style length - pollinator dependence axis (PC3) for number of visits and specialization. On PC1, number of visits was higher for plant species with more flowers but was lower for plant species with larger flowers (figure S10a). However, specialization showed the opposite trend on this axis (figure S10g). On PC3, the number of visits was lower for plants with shorter styles and higher pollinator dependence but higher for plant species with longer styles and lower pollinator dependence (figure S10c). Again, specialization showed the opposite trend to number of visits (figure S10i).
